Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2024-05-28 report/appendix/l2_constr.Rmd
Submitted paper

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals. The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          placement = "b",
          stub_indents = list("Hesitation showdown" = 1:3,
                              "Hesitation probability" = 4:6) )

>>>>

apa_table(tmp[,-1],
          align = c("l", "r", "r", "r", "r"), 
          caption = "Mixture model estimates for language effect. Cell means are shown the slowdown for hesitant interkey intervals when writing in L1 and L2 and the probability of hesitant keyintervals. The language difference is shown on log scale (for durations) and logit scale for probability of hesitant key intervals. 95% PIs in brackets.",
          note = "PIs are probability intervals. BF is the evidence in favour of the alternative hypothesis over the null hypothesis.",
          font_size = "small",
          stub_indents = list("Hesitation showdown" = 1:3,
                              "Hesitation probability" = 4:6) )

### 2024-05-28 report/appendix/models.Rmd
Submitted paper

This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\text{log}\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.

>>>>

This model can be described as in equation \ref{eq:unimodloggaus} in which the distribution $\mathcal{N}()$ was replaced by $\text{log}\mathcal{N}()$.

### 2024-05-28 report/appendix/postparam_constr.Rmd
Submitted paper

beta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~beta_plots[[.]] + theme(strip.text = element_blank()))
delta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~delta_plots[[.]] + theme(strip.text = element_blank()))
theta_plots[c(2,3,5,6)] <- map(c(2,3,5,6), ~theta_plots[[.]] + theme(strip.text = element_blank()))

@@

plot_fin_1 <- plot_param[[1]] / plot_param[[2]] / plot_param[[3]]
plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("../figures/psplots1constr.pdf", plot_fin_1, height = 6, width = 6)

>>>>

beta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~beta_plots[[.]] + theme(strip.text = element_blank()))
delta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~delta_plots[[.]] + theme(strip.text = element_blank()))
theta_plots[c(1,3,5,6)] <- map(c(1,3,5,6), ~theta_plots[[.]] + theme(strip.text = element_blank()))

@@

plot_fin_1 <-  plot_param[[2]] / plot_param[[1]] / plot_param[[3]]
plot_fin_1 <- grid.arrange(patchworkGrob(plot_fin_1), 
                           left = "Posterior estimate with 95% PIs",
                           bottom = "Transition location")
ggsave("../figures/psplots1constr.pdf", plot_fin_1, height = 6, width = 6)

### 2024-05-28 report/manuscript.Rmd
Submitted paper

---
title             : "Ideas cascade in parallel into keystrokes -- modelling writing hesitations as finite mixture process"
shorttitle        : "Ideas cascade in parallel into keystrokes"

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            : 
  papaja::apa6_pdf:
    keep_tex: FALSE

author: 
  - name          : "Jens Roeser"
    affiliation   : "1"
    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
    corresponding : yes 
    email         : "jens.roeser@ntu.ac.uk"
  - name          : "Mark Torrance"
    affiliation   : "1"
  - name          : "Rianne Conijn"
    affiliation   : "2"
  - name          : "Gunn Helen Ofstad"
    affiliation   : "3"
  - name          : "Evgeny Chukharev"
    affiliation   : "4"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
  - id            : "2"
    institution   : "Artificial Intelligence Systems Institute, Eindhoven University of Technology, The Netherlands"
  - id            : "3"
    institution   : "Educational Sciences and Humanities, University of Stavanger, Norway"
  - id            : "4"
    institution   : "Department of English, Iowa State University, Iowa"

abstract: "Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen. In contrast, semi-parallel models of writing assume that planning is not complete at production onset and operates in parallel with execution. Across six datasets from participants composing multi-sentence texts we instantiated these assumptions in two sets of Bayesian mixed effects models: (1) single-distribution models as assumed by the serial processing account, and (2) a finite (two-distribution) mixture model as assumed by the parallel processing account. We analysed interkey intervals at before-sentence, before word, and within word transitions. Model comparisons demonstrated strong evidence in favour of the parallel view across all datasets. When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning. However, we found -- across populations -- that interkey intervals at word and even at sentence boundaries were often too rapid to permit planning of what was written next. Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing."

keywords: "Cascading semi-parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"

>>>>

---
title             : "Typing in tandem: finite mixture models reveal fundamental parallelism in multi-sentence text composition"
shorttitle        : "Typing in tandem"

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE

author: 
  - name          : "Jens Roeser"
    affiliation   : "1"
    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
    corresponding : yes 
    email         : "jens.roeser@ntu.ac.uk"
  - name          : "Rianne Conijn"
    affiliation   : "2"
  - name          : "Evgeny Chukharev"
    affiliation   : "3"
  - name          : "Gunn Helen Ofstad"
    affiliation   : "4"
  - name          : "Mark Torrance"
    affiliation   : "1"

affiliation:
  - id            : "1"
    institution   : "Department of Psychology, Nottingham Trent University, United Kingdom"
  - id            : "2"
    institution   : "Human-Technology Interaction Group, Eindhoven University of Technology, The Netherlands"
  - id            : "3"
    institution   : "Department of English, Iowa State University, Iowa"
  - id            : "4"
    institution   : "Educational Sciences and Humanities, University of Stavanger, Norway"

# We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues. This work was supported by the National Science Foundation under Grant No. 2016868: "ProWrite: Biometric feedback for improving college students' writing processes." and UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production difficulties shape text formation."
# 
# Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024). For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see https://rpubs.com/jensroes/mixture-models-tutorial.

abstract: 'Classical serial models view the process of producing a text as a chain of discrete pauses, during which the next span of text is planned, and bursts of activity during which this text is output onto the page or computer screen. In contrast, semi-parallel models of writing assume that planning is not complete at production onset and operates in parallel with execution. Across six datasets from participants composing multi-sentence texts we instantiated these assumptions in two sets of Bayesian mixed effects models: (1) single-distribution models as assumed by the serial processing account, and (2) a finite (two-distribution) mixture model as assumed by the parallel processing account. We analysed interkey intervals at before-sentence, before word, and within word transitions. Model comparisons demonstrated strong evidence in favour of the parallel view across all datasets. When pausing occurred, sentence initial pauses were longer than word initial pauses which is consistent with the idea that larger linguistic edges are associated with higher level planning. However, we found -- across populations -- that interkey intervals at word and even at sentence boundaries were often too rapid to permit planning of what was written next. Our results cannot be explained by the serial processing but are in line with the semi-parallel view of writing.'

keywords: "Cascade; parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"

### 2024-05-28 report/sections/conclusion.Rmd
Submitted paper

In contrast to the serial pause-and-burst view, the parallel view emphasises that writers do not necessarily pause to plan the upcoming language unit but plan in parallel to writing execution. Using the Bayesian framework we implemented models of both the serial model of writing and the parallel view. Model comparisons and the inspection of posterior parameter estimates support the parallel view of writing but provide strong evidence against the serial view. We found that interkey intervals across word and even sentence boundaries are frequently fluent and not interrupted but planning of upcoming linguistic information. This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g. young / L2 writers, students) and writing tasks (e.g. essays, syntheses).

>>>>

The parallel view of writing, in contrast to the serial view, claims that writers do not necessarily pause to plan the upcoming language unit but plan in parallel with output. In this paper we have demonstrated, across six datasets, that statistical models premised on a parallel-processing understanding of planning in written composition outperform models premised on serial processing. We also found that interkey intervals across word and even sentence boundaries are frequently at timescales that are not consistent with planning of upcoming linguistic information. This pattern was found to be largely consistent for populations with different levels of writing experience and languages (e.g. young / L2 writers, students) and across different composition tasks. Our argument that written composition -- a very common form of language production -- is a largely parallel process is probably uncontroversial. However, the findings we presented in this paper constitute the first robust evidence in support of this claim.

### 2024-05-28 report/sections/discussion.Rmd
Submitted paper

The classical serial view [@flower1980ynamics] characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles. Text is therefore produced as a series of production bursts bounded by pauses. The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward]. The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is the pause duration doesn’t necessarily correlated with planning necessary for next output. We implemented  these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing. Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view. In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.

@@

This is what we found. At all linguistic edges we found a distribution of interkey intervals with durations that, in previous research, have been attributed just to motor execution (refs). For mid-word interkey intervals, as might be expected, the mixture model predicted that the vast majority of interkey intervals fall within this distribution. Across the six datasets that we sampled, the central tendency for the duration of mid-word interkey intervals was broadly similar to mean mid-word interkey intervals observed in previous studies. Importantly, however, these short interkey intervals were also present at higher-order text edges. Across all datasets we found, at minimum, a non-trivial minority of sentence-initial interkey intervals were sampled from the short distribution (i.e. with durations that could only be associated with motor execution). In two cases -- for adult writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution. Thus, although as might be expected there was a considerably increased tendency for longer interkey intervals at higher order boundaries: Planning associated with planning a new sentence is, probabilistically, much more likely to spill beyond output of the previous sentence and therefore affect duration of the sentence-initial interkey interval. However the presence of a non-trivial proportion of interkey intervals with a duration that is inconsistent with anything other than motor execution can only be explained with recourse to theories that permit preparation in parallel with preceding output.

@@

We observed two more differences across datasets -- other than the patterns summarised above for C2L1 -- that are worth highlighting. Both can largely be explained on the basis of differences associated with the population (writing competence / experience) that these samples were obtained from and text genre they were asked to produce [see also @conijn2019understanding]: The first exception concern the GE2 and SPL2 datasets: pauses were more likely to occur before sentences than before words; this difference was absent in the remaining data sets. **TODO: I don't have an explanation for this.** Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production. (2) pause durations did not vary across transition locations in the LIFT dataset. The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning content and text structure: the structure is often kept similar to the source in synthesis writing [“structural isomorphism”’, @hidi1986producing]. However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents or generating new sentence structures.

An observation that was not strictly predicted by the parallel view is that the unconstrained mixture model M5 that allowed the distribution over durations of fluent transitions to vary between transition locations showed a higher predictive performance. This can be explained on the basis of low-level keyboarding activity: It is possible that letter bigrams are systematically produced faster than transitions between space and a subsequent letter -- as in before-word transitions -- or shift -- at before sentence locations. This is for two reasons: first, and importantly, highly frequent bi- and even trigrams may be stored, retrieved and executed as one motor code [@behmer2016crunching; @grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants] because of the strong association between letters. This association reduces the length of transition durations. It is less likely that the same association exists for space-letter and space-shift pairs. Second, in two datasets (PLanTra, LIFT) before-sentence transitions are the sum of the transitions between space, shift and the following character key and therefore necessarily longer than other transitions. We found no evidence for the latter possibility (see Results section and Appendix \ref{key-combination-effect}).

A possible concern with our result -- that two-distributions mixture models showed substantially better predictive performance -- is that as mixture models have more parameters they might always outperform single-distribution models. This concern was preempted by using cross-validation techniques for model comparison which is preventing model overfit. In addition we used a simulation approach to compare a single distribution model and a two distributions mixture model. We simulated two datasets that were either based on a single distribution data-generating process or a process with a two-distributions process. This approach allows us to test the predictive performance of models in a context where we know the true underlying data generating process -- a single distribution vs a mixture of two distributions -- and we can test to what extent these models can successfully uncover the true parameter values. Importantly, we can test the performance of a mixture model for data that come from a single process. Modelling details can be found in Appendix \ref{simulation}. As might be expected, we found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process. Parameters were not successfully uncovered when we switched model type and dataset. Importantly though, cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data simulated from a two-distributions mixture process, but not when applied to data based on a single-distribution process. Therefore we can rule out the possibility that the statistical support for the two-distributions mixture-models can be explained on the basis of their larger number of parameters (i.e.model overfitting).

Throughout this paper we assumed that hesitations reflect mental planning of upcoming ideas and encoding of linguistic units. Alternatively, at least some if not all pauses in the writing process might be the result of reading or looking back into previously produced text. There is limited research that used eye tracking to investigate regressive gaze behaviour during writing [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading]. There are at least four reasons why eye movements away from the inscription point might occur during writing: First, reading is necessary for revision (i.e. writers must check if writing goals have been met). Second, writers might look for or correct spelling or grammatical errors. We removed transitions that terminated in an editing operation so the pauses we detected are unlikely to reflect revisions of the previous produced text. Third, writers might look at the source they are writing from. For two datasets (PLanTra, LIFT) which involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.^[As for PlanTra, most participants copy-pasted the source text into their word log. In other words, pauses might to some extent result from reading the source text.] Fourth, writers might use text to cue or reinstate ideas after interruption of the writing process [@johansson2010looking; @hayes2013new].

As for the fourth reason for why writers might look back in their text: frequencies of lookbacks during writing follow a similar pattern as the one we observed for pausing: for example @torrance2016reading reported that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]. In contrast to this observation -- that lookbacks are more frequent before sentences compared to words -- we found that pauses appear equally often before sentences and words under our two-distributions mixture-model. Finally, even though reading is likely to explain some proportion of our results, lookbacks in previously written text are overall rare, certainly in samples of university students [@torrance2016reading; @chukharev2019combined]. Also, pause frequency did not change when the produced text was masked (see comparisons for the CATO and GE2 data in Appendix \ref{masking-effect-cato-ge2}), and therefore if reading previously produced text was not possible.

A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidates for lower pause bound to handle the complex distribution of latency from spontaneous text production. Such thresholds are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent]. There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06;@van2016keystroke], i.e. a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke]. Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause [see e.g. @wen06] they would be arbitrary. For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word. Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision. So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs in the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs. However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword]. Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying].

>>>>

The classical serial view [@flower1980ynamics] characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles. Text is therefore produced as a series of production bursts bounded by pauses. The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward]. The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn’t necessarily correlate with planning necessary for next output. We implemented  these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing. Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view. In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.

@@

This is what we found. At all linguistic edges we found a distribution of interkey intervals at durations to motor execution at durations substantially below a threshold that might plausibly be necessary for, for example, lexical / orthographic retrieval [@grudin1982digraph; @ostry1983determinants; @terzuolo1980determinants; @waes2019]. For mid-word interkey intervals, as might be expected, the mixture model predicted that the vast majority of interkey intervals fall within this distribution. Across the six datasets that we sampled, the central tendency for the duration of mid-word interkey intervals was broadly similar to mean mid-word interkey intervals observed in previous studies. Importantly, however, these short interkey intervals were also present at higher-order text edges. Across all datasets we found, at minimum, a non-trivial minority of sentence-initial interkey intervals were sampled from the short distribution (i.e. with durations that could only be associated with motor execution). In two cases -- for writers sampled in the PLanTra and Lift datasets -- the majority of interkey intervals fell within this distribution. Thus, although as might be expected there was a considerably increased tendency for longer interkey intervals at higher order boundaries: Planning associated with planning a new sentence is, probabilistically, much more likely to spill beyond output of the previous sentence and therefore affect duration of the sentence-initial interkey interval. However the presence of a non-trivial proportion of interkey intervals with a duration that is inconsistent with anything other than motor execution can only be explained with recourse to theories that permit preparation in parallel with preceding output.

@@

We observed two more differences among datasets that are worth highlighting. Writers sampled in the GE2 and SPL2 datasets were more likely to pause before sentences than before words [see also @medimorec2017pauses; @wen02]. In the remaining datasets from writers who were either younger (and not educationally-selected) or writing in second language,  hesitation was equally probable pre-word and pre-sentence. This is most readily attributed simply to increased demands of lexical and / or orthographic processing in these samples.

 Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production. (2) pause durations did not vary across transition locations in the LIFT dataset. The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis [“structural isomorphism”, @hidi1986producing]. However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.

 We are, therefore, interpreting the superior predictive performance of the two-distribution mixture model as strong evidence for parallelism in written composition. An alternative, more prosaic explanation might be that two-distribution models outperform single distributions simply because they estimate more parameters; two distributions will always provide better fit than one distribution models due just to overfitting. There are two related reasons not to prefer this explanation. First, the cross-validation methods that we used to compare model predictive performance favour more parsimonious models and are therefore robust to overfitting. Second, in the analysis described in Appendix \ref{simulation} we directly tested the hypothesis that modelling data randomly sampled from a single log-normal distribution as a two-distribution mixture will provide better fit than modelling as a single distribution. We found that the two-distribution mixture model was successful at uncovering the parameter values of the data generated with a two-distributions mixture process; we observed the same for the single-distribution model applied to the data generated with a single distribution process. Parameters were not successfully uncovered when we switched model type and dataset. Cross-validation rendered a substantially higher predictive performance for the mixture model compared to the single-distribution model for data sampled from a two-distributions mixture process, but not when applied to data based on a single-distribution process. Therefore we can rule out the possibility that the statistical support for the two-distributions mixture-models can be explained on the basis of their larger number of parameters (i.e.model overfitting).

Throughout this paper we have been attributing longer (second-distribution) interkey intervals to spill-over from upstream pre-motor processing that could not be completed in its entirety in parallel with previous output. These longer inter-key intervals may also, on occasion, be associated with activity that is initiated after the preceding process and is entirely bounded by the duration of the pause. We are thinking here specifically of the tendency, particularly at higher order text boundaries, for writers to glance back into their existing text [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading]. @torrance2016reading found that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]. There are broadly two reasons why writers might look back into their existing text.^[Eye movement between key presses in writing-from-sources tasks will also be associated with reading source texts. Note, however, that for  two datasets (PLanTra, LIFT) that involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.]  Lookback may be associated with monitoring just-produced text for errors. In the present study we excluded interkey intervals that preceded deletion or cursor movement, and also intervals at timescales that might be associated with strategic reviewing of large spans of text (e.g., reading text-already-written from top to bottom). However it may be that some longer intervals were associated with monitoring that failed to find errors that needed correcting. A second function for lookback may be to support planning next-text [@johansson2010looking; @hayes2013new; @torrance2016reading]. The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next. This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this “temporal friction” [@van1991handwriting] becomes overloaded. Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}). This finding is consistent both with lookback being only a minor contributor to interkey interval and, where it does occur, serving a rapid memory-refresh function. There is also some evidence, at least in the context of handwritten production, that eye movements back into already-written text themselves occasionally occur in parallel with output [@Alamargot2007].

The findings we report in this paper are, we argue, the first robust evidence for parallelism in written composition of multi-sentence texts. The conclusion that text production processes are  fundamentally parallel will not come as a surprise to most researchers. Parallel processing has a long history in theories of speech production [@gar75; @levelt1989speaking; @dell1986; @chang2006; @dell1999connectionist; @dell1992stages]. The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation [@christiansen2016now]. Spoken production is “just-in-time”. Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed [@flower1980ynamics]. Our findings suggest that this is not the case. The language production processes that underlie text production follow the same just-in-time principle as speech.

<!-- A practical advantage of the mixture-model approach [also @roeser2021modelling] is that we are not required to stipulate interval thresholds as plausible candidates for lower pause bound to handle the complex distribution of latency from spontaneous text production. Such thresholds are widely used by writing researchers, a strategy that was inherited from early research in speech production [for review see @rochester1973significance], with studies dating back to at least the mid 1990s [@foulin1998extent]. There is one central problems with this approach: A threshold requires a definition of what passes as a pause [@wen06;@van2016keystroke], i.e. a pause criterion threshold often set to 2 secs [@chanquoy1996writing; @kaufer1986composing; @sullivan2002self; @wen02] or some other lower bound [@chukharev2014pauses; @connelly2012predicting; @leijten2013keystroke]. Even if these thresholds were adjusted relative to factors such as writing medium, experience of writer, and text location of pause [see e.g. @wen06] they would be arbitrary. For example, a sentence-initial pause of 2 secs has a very different interpretation from a 2 secs pause that occurs before or within a word. Our current understanding of the processes that underlie text production does not provide a strong theoretical basis on which to make this decision. So while a 2 secs threshold undoubtedly captures an interesting distinction -- processing that occurs in the range zero to 2 secs is very likely to be qualitatively different from processing that takes more than 2 secs. However the same could be argued for any threshold between perhaps 250 msecs and 10 secs that a researcher might care to choose [@chenu2014interword]. Mixture models provide a principled statistical framework that allows the researcher to model behavioural data that come from a combination of cognitive processes without imposing threshold values [see also @almond2012preliminary; @baaijen2012keystroke; @hall2022constructing; @li2021identifying]. -->

### 2024-05-28 report/sections/introduction.Rmd
Submitted paper

This semi-parallel cascaded account of the composition process therefore gives a rather different understanding of inter-keystroke intervals. Interkey intervals result from one of two sets of underlying mental processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then interkey intervals are determined just by the time needed for executing finger movements (i.e. just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly, then interkey intervals are determined just by time taken to complete upstream processes and not on finger movement. These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke. [@roeser2021modelling]. The parallel account provides an explanation for why writers sometimes hesitate before sentence (or word) onset, but often do not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

@@

The fourth model was premised on the assumption that the processing that results in a keystroke runs, in part, in parallel with preceding output. As we have discussed, this model assumes two data-generating processes. We therefore followed the finite (two-distribution) mixture-modelling approach described by@roeser2021modelling. This model assumes that interkey intervals result from a mixture of: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels that cause delays in this information flow. Therefore interkey intervals are not directly and sufficiently determined by transition location -- as in the serial account -- but are also associated with different probabilities that a hesitation occurs.

```{r eval = F}
#Hesitations in writing have potential disadvantages. Writers can, as we have noted, pause at any point without this affecting the eventual communicative effect of their text. However language processing is subject to what @christiansen2016now describe as a fundamental "now-or-never" bottleneck. Buffering is transient and as a consequence of written (and spoken) production are just-in-time systems: Production must flow down the cascade of processes from message to finger movements without significant interruptions. If there is delay -- if, for example, a writer struggles to retrieve the right word or its spelling -- there is a risk that output from upstream processes, and particularly the chain of ideas that the writer wanted to communicate, will be lost. @ronneberg2022process coined this possibility the "process-disruption hypothesis".
```

>>>>

This semi-parallel cascaded account of the composition process therefore gives a rather different understanding of inter-keystroke intervals. Interkey intervals result from one of two sets of underlying mental processes. When upstream processes output at a rate equal to or faster than can be used for finger movement planning, then interkey intervals are determined just by the time needed for executing finger movements (i.e. just by processing below the last buffer in the processing cascade). However if one or more upstream processes provide output more slowly, then interkey intervals are determined just by time taken to complete upstream processes and not on finger movement. These two distinct data-generating processes mean that interkey intervals are a mixture of two distributions, one associated with the very rapid processing necessary for actioning keystrokes, and another that forms as a result of delays resulting from some combination of processing at the semantic, syntactic, lexical, or orthographic level not being complete at the time of the immediately preceding keystroke [@roeser2021modelling]. The parallel account provides an explanation for why writers sometimes hesitate before sentence (or word) onset, but often do not. This is not because in some cases sentences are not planned, or planning was postponed until after sentence onset, but that this planning was completed in parallel with previous output.

@@

The fourth model was premised on the assumption that the processing that results in a keystroke runs, in part, in parallel with preceding output. As we have discussed, this model assumes two data-generating processes. We therefore followed the finite (two-distribution) mixture-modelling approach described by @roeser2021modelling. This model assumes that interkey intervals result from a mixture of: (1) uninhibited activation flow into motor programmes and (2) interruptions at higher levels that cause delays in this information flow. Therefore interkey intervals are not directly and sufficiently determined by transition location -- as in the serial account -- but are also associated with different probabilities that a hesitation occurs.
